---
title: "Class_03 | In-Class Assignment | Intro to R"
author: "Sean Mussenden"
date: "9/24/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, paged.print=TRUE)
```

## Objective

The purpose of this in-class assignment is to build on the information you learned in last week's in-class lab:

* Writing R code for data analysis and exploration in the R Studio environment, using R projects (.Rproj) and R markdown files (.Rmd).  
* Installing and loading packages, including the Tidyverse framework of packages. 
* Loading data with functions from Readr, the Tidyverse package for loading data. 
* Getting a sense of data with functions like glimpse(), summary(), and View().
* How to examine a subset of the data, using functions from Dplyr, the Tidyverse package for wrangling data. 
* How to sort data, using a function from Dplyr, the Tidyverse package for wrangling data. 
* Working with data on opioid shipments in the U.S.

## Tasks, Turning it In, Getting Help

At several points throughout this document, you will see the word **Task**.  

This indicates that you need to do something, generally creating a code block and writing custom code.  

When you are finished, you should save your R markdown file and Knit it as an HTML file.  Upload both the R Markdown file and HTML file to ELMS. 

Need help?  You are welcome to do the following things:

* Refer to the previous week's lab. 
* Use Google or search Stack Overflow. Try searching for your error message or translating your problem into basic terms. 
* Check out the excellent [R for Data Science](https://r4ds.had.co.nz/index.html)
* Take a look at the [Cheatsheets](https://www.rstudio.com/resources/cheatsheets/) and [Tidyverse documentation](https://www.tidyverse.org/).
  * [RStudio cheat sheet](https://www.rstudio.com/resources/cheatsheets/#ide)
  * [Readr cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf) and [Readr documentation](https://readr.tidyverse.org/).
  * [Dplyr cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) and [Dplyr documentation](https://dplyr.tidyverse.org/)
* After you've spent 5 minutes trying to solve the problem on your own, ask your neighbor and if they don't know, ask me! 

## Setup

Take the following steps to set up your document:

1. Download the ZIP file and open the folder on your desktop. It should contain this document, class_03.Rmd, and a data folder with one TSV. 
2. Open this file in RStudio. 
3. Rename this file "class_03_FIRSTNAME_LASTNAME.Rmd". 
4. Create a new R project inside of this folder, which will set the working directory in this folder.   

## Load Packages

Generally, it's good practice to load packages you wish to work with at the beginning of your R markdown file.  

**Task**: In the code block below, write and execute the simple function to load the Tidyverse family of packages.  

```{r}
# Load the Tidyverse family of packages below. 

# Remember: if this is the first time you've used this package on this computer, you'll have to install it first with install.packages()  

# Write the function to load the Tidyverse below here. 

library(tidyverse)

```

## Load Data 

For this exercise, we will be working with a small subset of the DEA's ARCOS database, which documented shipments of 76 billion opioid pills between 2006 and 2012, during the peak of the opioid epidemic. We will be working with a subset of shipments to Logan County, West Virginia, which was flooded with hydrocodone and oxycodone during that period.  It received more pills per capita than all but a handful of other counties in the U.S. 

The data was obtained after a lengthy legal battle by the Washington Post and the Charleston Gazette-Mail, and released by the Washington Post in raw and aggregated form. [Washington Post "Digging into the DEA's pain pill database" page](https://www.washingtonpost.com/graphics/2019/investigations/dea-pain-pill-database/). 

There is one record (row or observation) per transaction (a shipment from a manufacturer to a distributor to a pharmacy. A data dictionary is available here: [ARCOS Registrant Handbook](https://www.deadiversion.usdoj.gov/arcos/handbook/full.pdf).

**Task**: Create a code block below, and write and execute the function to load in the data of shipments to Logan County, West Virginia and store it as an object called logan. Write a comment describing what you are doing.  

``` {r}

# Read in the tsv file and assign it to object 'logan'
logan <- read_tsv("data/arcos-wv-logan-54045-itemized.tsv")

```

## Examine the Data

Now that the data is in, spend some time examining it. 

**Task**: Create a code block below, and write and execute a function to open and view the spreadsheet in a new window tab that you can examine.  In that browser window, use your mouse to sort by a handful of columns, highest to lowest and lowest to highest.  In that browser window, practice turning on filters with your mouse to examine a subset of the data.

``` {r}

# View logan spreadsheet
#View(logan)

```

**Task**: Create a code block below and use the glimpse function to examine the shape of the data. Answer each of the following questions in a comment inside the code block:
* How many rows does the data have? How many columns?   
* Each column is stored in one of two formats: "chr", or strings of characters, generally used for text; and "dbl", a number format on which mathematical operations can be performed.  Do you see any examples of a column stored as a "dbl" that doesn't really need to be, because we would never do math on it? Are there any columns stored as character that you think should be stored in another format entirely? 
* Based on the information here, how are missing values represented in this data set?

``` {r}

# Glimpse logan
glimpse(logan)

# The data has 65,359 rows and 42 columns.

# Fields stored as <dbl> that don't need to be: REPORTER_ZIP, BUYER_ZIP, DRUG_CODE, TRANSACTION_ID

# Fields stored as <chr> that should be stored as another format: Nothing particularly jumps out. If there were multiple product types in a shipment, it might be useful to use <list> instead of <chr> for Product_Name or Ingredient_Name.

# Missing values appear to be represented as "null."

```

**Task**: Create a code block below and use the summary function to examine the shape of the data. Answer each of the following questions in a comment inside the code block:
* What does BUYER_ZIP represent? What is the mean value for BUYER_ZIP?  Is the mean value a meaningful thing for us to use? Why or why not? 
* What does DOSAGE_UNIT represent? What are the min, max and mean values for DOSAGE_UNIT? Are these meaningful statistics? Why or why not?

```{r}

# Summarize logan
summary(logan)

# BUYER_ZIP represents the zip code of the pharmacy that bought the pills. The mean value, which is 25594, is not useful to us at all because the zip code is a representation of location, not a mathematical factor that we would want to examine further.

# DOSAGE_UNIT represent the number of pills in each shipment. The min, max, and mean values are 10, 25000 and 697.5, respectively, and they are meaningful statistics because they help us understand the nature of the shipments made to Logan County.

```

## Analysis - Selecting

This data set is wide, with 42 columns that may not always be useful for questions we want to ask.  Let's use a function to reduce the n umber of columns. 

**Task**: Suppose we want to create a subset of the data that allows us to look at the manufacturer-distributor-pharmacy pipeline for each shipment.  Update the code block below and use a function to return only the following four columns in order for each record: the name of the manufacturer, the name of the distributor and the name of the pharmacy, followed by the number of pills in that record. Store this as an object called logan_pipeline.  Then write a function to print out the first 1,000 rows in the R Markdown document below, and another function to view the returned spreadsheet in a new window.  Write comments inside the code block to explain what you're doing. After you have examined the sheet, comment out the function you wrote to view it in a new window.

```{r rows.print=1000}

# Get name of manufacturer, distributor, pharmacy and dosage unit and store it as an object called logan_pipeline
logan_pipeline <- logan %>% select(Combined_Labeler_Name, REPORTER_NAME, BUYER_NAME, DOSAGE_UNIT)

# Look at it, popping it out in a new window
#View(logan_pipeline)

# Also print it in our R markdown file below our codeblock (only the first 1K rows returned)
print(logan_pipeline)

```

**Task**: Let's trim our data set just a bit. Create a code block below and use a function to return every column EXCEPT for the first four columns. Store this as an object called logan_all_but_four.  Then write a function to print out the first 1,000 rows in the R Markdown document below, and another function to view the returned spreadsheet in a new window. Write comments inside the code block to explain what you're doing. After you have examined the sheet, comment out the function you wrote to view it in a new window.

```{r rows.print=1000}

# Build new object 'logan_all_but_four' - found formula at https://www.listendata.com/2015/06/r-keep-drop-columns-from-data-frame.html
logan_all_but_four <- logan %>% select(-1:-4)

# View logan_all_but_four
#View(logan_all_but_four)

# Print logan_all_but_four
print(logan_all_but_four)

```

**Task**: Let's just look at the pharmacy information. Create a code block below and use a function to return every column that contains the word "BUYER", and none of the other columns. Store this as an object called logan_buyer.  DO NOT just write out every single column inside of your function with the word "BUYER". Find a function that will allow you to search within the column names for the word instead. Stuck? Look at the Dplyr documentation or cheat sheet.  Or try and craft a generalized search query for Google that might find you the right answer: "select all columns contain a word dplyr". Write a function to print out the first 1,000 rows in the R Markdown document below, and another function to view the returned spreadsheet in a new window. Write comments inside the code block to explain what you're doing. After you have examined the sheet, comment out the function you wrote to view it in a new window.

```{r rows.print=1000}

# Build new object 'logan_buyer' - found function in Dplyr documentation
logan_buyer <- logan %>% select(contains("BUYER"))

# View logan_buyer
#View(logan_buyer)

# Print logan_buyer
print(logan_buyer)

```

## Removing Objects

**Task**: Our workspace is getting a little bloated. Create a code block below and write a single function to remove the following four objects from the environment: logan_pipeline, logan_all_but_four, logan_buyer. Even better, can you write a single function that keeps the logan object but removes everything else? Try searching: "remove all but one object in R".  In the code block, write comments explaining what you're doing.

```{r}

# Remove all objects from environment besides logan - found solution at https://stackoverflow.com/questions/6190051/how-can-i-remove-all-objects-but-one-from-the-workspace-in-r
rm(list=(ls()[ls()!="logan"]))

```

## Analysis - Selecting and Sorting

**Task**: Let's sort to organize our data a bit. Write a single function that a) returns only the names of the manufacturer, the distributor and the pharmacy and the number of pills AND THEN sorts from A to Z by manufacturer name. Store it as an object called logan_pipeline_sorted. Then write a function to print out the first 1,000 rows in the R Markdown document below, and another function to view the returned spreadsheet in a new window. Write comments inside the code block to explain what you're doing and what you found. After you have examined the sheet, comment out the function you wrote to view it in a new window.   

```{r print.rows=1000}

# Build 'logan_pipeline_sorted' with only names of mfr, distro, pharmacy, and no. of pills; sort A-Z by manufacturer
logan_pipeline_sorted <- logan %>% select(Combined_Labeler_Name, REPORTER_NAME, BUYER_NAME, DOSAGE_UNIT) %>% arrange(Combined_Labeler_Name)

# View logan_pipeline_sorted
#View(logan_pipeline_sorted)

# Print first 1,000 rows of logan-pipeline-sorted
print(logan_pipeline_sorted)

```

**Task**: Now let's sort in the other direction. Write a single function that a) returns only the names of the manufacturer, the distributor and the pharmacy and the number of pills AND THEN sorts Z to A by pharmacy name.  Store it as an object called logan_pipeline_sorted, overwriting the object you created in the last example. Then write a function to print out the first 1,000 rows in the R Markdown document below, and another function to view the returned spreadsheet in a new window. Write comments inside the code block to explain what you're doing and what you found. After you have examined the sheet, comment out the function you wrote to view it in a new window.


```{r print.rows=1000}

# Build 'logan_pipeline_sorted' with only names of mfr, distro, pharmacy, and no. of pills; sort Z-A by pharmacy
logan_pipeline_sorted <- logan %>% select(Combined_Labeler_Name, REPORTER_NAME, BUYER_NAME, DOSAGE_UNIT) %>% arrange(desc(BUYER_NAME))

# View logan_pipeline_sorted
#View(logan_pipeline_sorted)

# Print first 1,000 rows of logan-pipeline-sorted
print(logan_pipeline_sorted)

```

## Analysis - Selecting and Sorting and Filtering

**Task**: Let's look at only the largest transactions in our data set.  Write a single function that a) returns only the names of the manufacturer, the distributor and the pharmacy and the number of pills AND THEN b) sorts from highest to lowest on number of pills AND THEN d) returns only those transactions greater than or equal to 15000. Because this query only returns a small subset of data (less than 1,000 records), don't worry about assigning it to an object or popping it out in a new window.  Just allow it to print below the code block.  Note: you won't have to write out a print() function to do this.

```{r}

# select names of mfr, distro, pharmacy, and no. of pills; sort hi-lo by no. of pills (transactions >= 15000 only)
select(logan, Combined_Labeler_Name, REPORTER_NAME, BUYER_NAME, DOSAGE_UNIT) %>% arrange(desc(DOSAGE_UNIT)) %>% filter(DOSAGE_UNIT >= 15000)

```

**Task**: Let's drill in on one distributor.  Replicate the same query as above, but this time also filter so only "MCKESSON CORPORATION" records are returned.

```{r}

# select names of mfr, distro, pharmacy, and no. of pills; sort hi-lo by no. of pills (transactions >= 15000 with McKesson as reporter only)
select(logan, Combined_Labeler_Name, REPORTER_NAME, BUYER_NAME, DOSAGE_UNIT) %>% arrange(desc(DOSAGE_UNIT)) %>% filter(DOSAGE_UNIT >= 15000, REPORTER_NAME == "MCKESSON CORPORATION")

```

**Task** Let's drill in even further. Replicate the last query, but this time add a filter that searches for the word "ARACOMA" in the pharmacy name column.

```{r}

# select names of mfr, distro, pharmacy, and no. of pills; sort hi-lo by no. of pills (transactions >= 15000 with McKesson as reporter and pharmacy containing "ARACOMA" only)
select(logan, Combined_Labeler_Name, REPORTER_NAME, BUYER_NAME, DOSAGE_UNIT) %>% arrange(desc(DOSAGE_UNIT)) %>% filter(DOSAGE_UNIT >= 15000, REPORTER_NAME == "MCKESSON CORPORATION", str_detect(BUYER_NAME, "ARACOMA"))

```

## Your own questions

**Task**: Create three separate code blocks below.  Use selecting, sorting and filtering to ask and answer three questions of your own design. Write out the question above each code block, and the answer you discovered writing the code. Use comments to explain what you're doing.  

Are shipments of 5,000 pills or larger more likely to be oxycodone or hydrocodone? Print only the top 100 shipments.

```{r print.rows=100}

# Return shipments >= 5000 pills, sorted hi-lo by dosage unit, and which drug was shipped
logan_5k <- logan %>% select(DOSAGE_UNIT, DRUG_NAME) %>% arrange(desc(DOSAGE_UNIT)) %>% filter(DOSAGE_UNIT >= 5000)

# Print logan_5k
print(logan_5k)

```

Which pharmacies are buying the most oxycodone pills?

```{r}

# Return pharmacy name, city, drug name and dosage unit, sorted hi-lo, filtered for oxycodone
logan_top_oxy <- logan %>% select(BUYER_NAME, BUYER_CITY, DRUG_NAME, DOSAGE_UNIT) %>% arrange(desc(DOSAGE_UNIT)) %>% filter(DRUG_NAME == "OXYCODONE")

# Print logan_top_oxy
print(logan_top_oxy)

```

Are individual practitioners buying hydrocodone or oxycodone? How much?

```{r}

# select buyer name, buyer type, drug name, dosage unit, sorted hi-lo by dosage, filtered for buyer type containing 'practitioner'
logan_practitioner <- logan %>% select(BUYER_NAME, BUYER_BUS_ACT, DRUG_NAME, DOSAGE_UNIT) %>% arrange(desc(DOSAGE_UNIT)) %>% filter(str_detect(BUYER_BUS_ACT, "PRACTITIONER"))

# print logan_practitioner
print(logan_practitioner)

```

## Output 

**Task**: Spell check your document in R Studio. Figure out how to do this by using the help search menu. Save your file.  Knit it to an HTML document, making sure it compiles.  Open it in a browser to be sure. 

```{r include=FALSE}
library(spelling)
spell_check_files("class_03_Sean_McGoey.Rmd")
```

## Submission

**Task**: On ELMS, you'll upload both the html file produced after you knit this document.  And you'll upload the R Markdown file. 
